---
output: github_document
---


# M2algorithmique Vignette
### Vincent Runge
#### LaMME, Evry Paris-Saclay University
### March 17, 2025


> [Package Presentation](#pp)

> [Quick Start](#qs)

> [Examples](#ex)

> [Analysis](#an)


___ 

<a id="pp"></a>

## Package Presentation  

The `M2algorithmique` R/Rcpp package is an **example package** designed to help students develop their own R/Rcpp packages as part of the **M2 Algorithmic courses in the Data Science master's program at Université d'Évry Paris-Saclay**.  

This package provides implementations of various algorithmic strategies in both R and Rcpp, including:

- **Sorting algorithms**: Insertion Sort and Heap Sort  
- **Shortest path algorithms**: Dijkstra's algorithm with naive, binary heap, and Fibonacci heap implementations  
- **Travelling Salesman Problem (TSP)**: Exact Branch and Bound algorithm, along with multiple heuristic approaches  

This package serves as a reference for students, demonstrating good practices in algorithm implementation and package development in Rcpp.

___ 

<a id="qs"></a>

## Quick Start

### Prerequisites for Package Development

To develop and use the package, install the necessary dependencies:

```{r, eval=FALSE}
install.packages(c("Rcpp", "RcppArmadillo", "devtools", "roxygen2", "testthat"))
```

### Installing the Package from GitHub

To install the M2algorithmique package, use:

```{r, eval=FALSE}
devtools::install_github("vrunge/M2algorithmique")
```

Then, load the package:

```{r}
library(M2algorithmique)
```

### Setting Up a C++ Compiler (Windows Users)

Since Rcpp requires a C++ compiler, Windows users must install Rtools:

Download and install Rtools from:
https://cran.r-project.org/bin/windows/Rtools/

For general R updates and additional package downloads, visit the CRAN:
https://cran.r-project.org/



### Package Structure  

- **One file = one function**: Each function is stored in a separate file for clarity. 

- **R and C++ function naming convention**:  
  - R functions (`fct`) are stored in the `R/` folder.  
  - Their corresponding C++ versions (`fct_Rcpp`) are in the `src/` folder.  
  - A **header file** (`.h`) can be used to define functions that are accessible 
    across multiple C++ files. In this package, we use for example **`TSP_auxiliary.h`** 
    to make the distance function globally available to all C++ implementations of heuristic TSP
    algorithms. To use this function in a C++ file, you need to include the header at the beginning of
    the file: `#include "TSP_auxiliary.h"`.

- **Documentation**:  
  - Help files for functions are available via `?fct`.  
  - Files in the `man/` folder are automatically generated using **Roxygen2**.  
  - To regenerate documentation when installing the package:  
    1. Click **Build** → **More** → **Configure Build Tools...**  
    2. Enable **Generate Documentation with Roxygen**  
    3. Click **Configure**, then check **Install and Restart** 
    
- **Key Package Files**:  
  - **`DESCRIPTION`**: Modify this file to personalize your package (e.g., email, description).  
  - **`NAMESPACE`**: Controls function exports. Customize it to expose only desired functions while keeping others internal.  
  - **`.Rbuildignore`**: Excludes files that are unnecessary for package building (e.g., the `Pour_les_etudiants/` folder).  
  - **`.gitignore`**: Prevents unnecessary files (e.g., `.o` object files) from being tracked on GitHub.  
  
- **Unit tests**. `tests/` folder and `testthat` integration:  
  - Unit tests are stored in the `tests/testthat/` folder.  
  - The package uses `testthat` to automate testing and ensure functions work correctly.  
  - To avoid warnings during `R CMD check`, `testthat` should be listed under **Suggests** in `DESCRIPTION`.  




___ 

<a id="ex"></a>

## Examples

### Sorting Algorithms (Recursive)


```{r}
n <- 10
v <- sample(n)
v
```


We implemeted 4 algorithms: 

- `insertion_sort` 
- `heap_sort` 
- `insertion_sort_Rcpp` 
- `heap_sort_Rcpp` 

They all have a unique argument: the unsorted vector `v`. Examples:

```{r}
insertion_sort(v)
heap_sort_Rcpp(v)
```




### Dijkstra Algorithms (Dynamic Programming)


### TSP Algorithms (Heuristic)  

We generate `n = 20` towns (villes) randomly in a **2D plane** using a uniform distribution:  
```{r}
n <- 20
villes <- matrix(runif(2 * n), n, 2)
```  

#### Visualization of the Towns  
```{r, echo=FALSE}
plot(villes[,1], villes[,2], xlim = c(0,1), ylim = c(0,1), xlab = "", ylab = "", asp=1, pty ="s")
```  

We implemented **four heuristic** algorithms and **one exact** algorithm for solving the **Travelling Salesman Problem (TSP)**:  

- **Heuristic Approaches:**  
  - `TSP_naif` (Naive approach)  
  - `TSP_cheapest` (Cheapest insertion heuristic)  
  - `TSP_nearest` (Nearest neighbor heuristic)  
  - `TSP_farthest` (Farthest insertion heuristic)  

- **Exact Algorithm:**  
  - `TSP_B_and_B` (Branch and Bound)  

Each algorithm also has an equivalent **Rcpp implementation** for performance optimization.  

#### Example

```{r, echo=FALSE}
res1 <- TSP_naif(villes)
res2 <- TSP_cheapest(villes)
res3 <- TSP_nearest(villes) 
res4 <- TSP_farthest(villes)
t1 <- tour_length(res1, villes)
t2 <- tour_length(res2, villes)
t3 <- tour_length(res3, villes)
t4 <- tour_length(res4, villes)
par(mfrow=c(2,2),
    oma = c(5,4,0,0) + 0.1,
    mar = c(1.5,1.5,1,0) + 0.1)
plot.TSP(tour = res1, data = villes, main = "naif", value = t1)
plot.TSP(tour = res2, data = villes, main = "cheapest", value = t2)
plot.TSP(tour = res3, data = villes, main = "nearest", value = t3)
plot.TSP(tour = res4, data = villes, main = "farthest", value = t4)
```


___

<a id="an"></a>

## Analysis  

The `Pour_les_etudiants` folder contains the following analysis files:  

- **`Sorting_analyse.Rmd`** – Analysis of sorting algorithms  
- **`Dijkstra_analyse.Rmd`** – Analysis of Dijkstra’s algorithm  
- **`TSP_analyse.Rmd`** – Analysis of the Traveling Salesman Problem (TSP)  

<span style="color:red">These documents, written in French, outline the type of analysis students are expected to conduct to validate this course.</span>

